The state of knowledge for marine invertebrate biodiversity in the continental US

### 2018-05-05 manuscript.Rmd
more text edits

---
title: "The state of knowledge for marine invertebrate biodiversity in the continental US"
date: ' ' #'`r format(Sys.time())`'
output:
  bookdown::pdf_document2:
    includes:
    #  before_body: ../template/doc_prefix.tex
      in_header: preamble.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc: no
fontsize: 12pt
link-citations: yes
csl: nature.csl
bibliography: /home/francois/Library/dark-plankton.bib
abstract: Museum collections have been recording our knowledge about the diversity and distribution of species for about 250 years. Until recently, this information was fragmented across institutions and difficult to access or synthesize. The digitization of biodiversity information is enabling the synthesis of collection data. Here we use publicly accessible digital biodiversity data to asses how well we know the marine invertebrate fauna of the continental United States. Globally, marine inverterbrates comprise about 75% of described marine species. We document `r format(n_distinct(combined_species$worms_valid_name), big.mark = ",")` species from the continental US waters in digital databases (OBIS and iDigBio), and explore spatial and temporal patterns of their distribution and collection. Databases do well at the fauna level, capturing ~xxx% of the species diversity. However, sampling bias obscure broad diversity patterns, such as differences between coasts and across latitude. Our knowledge about species richness and geographic range is limited by sampling effort, with `r prop_10_specimens_species`% of species known from $\leq$10 samples. While recent developments in high-throughput sequencing promise to greatly speed up documentation of marine biodiversity, interpretation of sequences requires anchoring results anchoring results to identified vouchers. DNA barcode sequences (COI) are presently available for only `r format_output(prop_spp_barcoded$p_barcoded * 100)`% of the US marine invertebrates, and many of these are from samples taken outside the US. Over a third of the barcoded species are complexes of of multiple cryptic lineages, indicating that the fauna is substantially richer than the above number indicate and needs to be taxonomically reevaluated as a whole. Substantially increased sampling, coordinated at the national scale, is needed to provide a meaningful baseline for marine biodiversity and for metabarcoding and metagenomics studies. This data will allow the detection of the impacts of climate change and other anthropogenic strains on marine biodiversity. Advances in computing, information systems, and sequencing technologies, have set the stage for a broad understanding of the dynamics of marine biodiversity, but this potentially is limited by the availablility of current samples, digitization, and taxonomic effort.
---

@@

The above estimates are based on current taxonomy, but many species as currently understood are complexes of multiple cryptic species [@Knowlton2000;@Meyer2005-evolution]. BOLD uses an algorithm to assess species-level genetic differentiation among samples, separating molecular operational units into different BINs (BOLD Index Numbers)[@Ratnasingham2013]. More than a third of the marine invertebrate species recorded from the US and represented in BOLD include multiple BINs (`r format_output(100 * stat_bold$p_spp_multi_bin)`% of `r format(stat_bold$n_spp_not_singleton, big.mark=",")` species). Given the relatively sparse DNA coverage among as well as within species, the actual level of crytpic differentiation is higher. Many of the BOLD sequences of US species are based on specimens collected outside the US. Thus both the diversity and species-level identity of the marine fauna needs to be comprehensively reevaluated.

Specimens identified under the same species but sampled from distant locations are especially prone to include cryptic species. While geographic information is missing for most records in BOLD (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them), of the `r format(bold_geography_stats$n_spp_with_records, big.mark=",")` marine invertebrates species from the US with BOLD records, only `r format_output(bold_geography_stats$p_within_eez * 100)`% have at least one record collected within the EEZ and `r format_output(bold_geography_stats$p_in_north_america * 100)`% list a North American country as the site of collection. It is crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced to capture their geographic range and  genetic diversity.

>>>>

---
title: "The state of knowledge for marine invertebrate biodiversity in the continental US"
date: ' ' #'`r format(Sys.time())`'
output:
  bookdown::pdf_document2:
    includes:
    #  before_body: ../template/doc_prefix.tex
      in_header: preamble.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc: no
fontsize: 12pt
link-citations: yes
csl: nature.csl
bibliography: /home/francois/Library/dark-plankton.bib
abstract: Museum collections have been recording our knowledge about the diversity and distribution of species for about 250 years. Until recently, this information was fragmented across institutions and difficult to access or synthesize. The digitization of biodiversity information is enabling the synthesis of collection data. Here we use publicly accessible digital biodiversity data to asses how well we know the marine invertebrate fauna of the continental United States. Globally, marine inverterbrates comprise about 75% of described marine species. We document `r format(n_distinct(combined_species$worms_valid_name), big.mark = ",")` species from the continental US waters in digital databases (OBIS and iDigBio), and explore spatial and temporal patterns of their distribution and collection. Databases do well at the fauna level, capturing ~xxx% of the species diversity. However, sampling bias obscure broad diversity patterns, such as differences between coasts and across latitude. Our knowledge about species richness and geographic range is limited by sampling effort, with `r prop_10_specimens_species`% of species known from $\leq$ 10 samples. While recent developments in high-throughput sequencing promise to greatly speed up documentation of marine biodiversity, interpretation of sequences requires anchoring results anchoring results to identified vouchers. DNA barcode sequences (COI) are presently available for only `r format_output(prop_spp_barcoded$p_barcoded * 100)`% of the US marine invertebrates, and many of these are from samples taken outside the US. Over a third of the barcoded species are complexes of of multiple cryptic lineages, indicating that the fauna is substantially richer than the above number indicate and needs to be taxonomically reevaluated as a whole. Substantially increased sampling, coordinated at the national scale, is needed to provide a meaningful baseline for marine biodiversity and for metabarcoding and metagenomics studies. This data will allow the detection of the impacts of climate change and other anthropogenic strains on marine biodiversity. Advances in computing, information systems, and sequencing technologies, have set the stage for a broad understanding of the dynamics of marine biodiversity, but this potentially is limited by the availablility of current samples, digitization, and taxonomic effort.
---

@@

The above estimates are based on current taxonomy, but many species as currently understood are complexes of multiple cryptic species [@Knowlton2000;@Meyer2005-evolution]. BOLD uses an algorithm to assess species-level genetic differentiation among samples, separating molecular operational units into different BINs (BOLD Index Numbers)[@Ratnasingham2013]. More than a third of the marine invertebrate species recorded from the US and represented in BOLD include multiple BINs (`r format_output(100 * stat_bold$p_spp_multi_bin)`% of `r format(stat_bold$n_spp_not_singleton, big.mark=",")` species represented by more than XX sequences). XX% of the fauna is only known from single barcodes at this stage. Given the relatively sparse DNA coverage among as well as within species, the actual level of crytpic differentiation is higher. Many of the BOLD sequences of US species are based on specimens collected outside the US. Thus both the diversity and species-level identity of the marine fauna needs to be comprehensively reevaluated.

Specimens identified under the same species but sampled from distant locations are especially prone to include cryptic species. While geographic information is missing for most records in BOLD (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records that match a species name in our checklist have no geographic coordinate information associated with them), of the `r format(bold_geography_stats$n_spp_with_records, big.mark=",")` marine invertebrates species from the US with BOLD records, only `r format_output(bold_geography_stats$p_within_eez * 100)`% have at least one record collected within the EEZ and `r format_output(bold_geography_stats$p_in_north_america * 100)`% list a North American country as the site of collection. It is crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced to capture their geographic range and  genetic diversity.
